---
title: "Prediction Based on Betting Markets"
output: pdf_document
---

Earlier in the chapter, we studied the prediction of election outcomes using polls.  Here, we study the prediction of election outcomes based on betting markets.  In particular, we analyze data for the 2008 and 2012 US presidential elections from the online betting company, called Intrade.  At Intrade, people trade contracts such as 'Obama to win the electoral votes of Florida.'  Each contract's market price fluctuates based on its sales.  Why might we expect betting markets like Intrade to accurately predict the outcomes of elections or of other events?  Some argue that the market can aggregate available information efficiently.  In this exercise, we will test this  *efficient market hypothesis* by analyzing the market prices of contracts for Democratic and Republican nominees' victories in each state.

The data files for 2008 and 2012 are available in CSV format as `intrade08.csv` and `intrade12.csv`, respectively. The variables in these datasets are:

--------------------------------------------------------------------------------
 Name                Description
 ------------------- ----------------------------------------------------------
 `day`               Date of the session 
 
 `statename`         Full name of each state (including District of 
                     Columbia in 2008) 
 
 `state`             Abbreviation of each state (including District of 
                     Columbia in 2008)
 
 `PriceD`            Closing price (predicted vote share) of Democratic 
                     Nominee's market
 
 `PriceR`            Closing price (predicted vote share) of Republican 
                     Nominee's market 
 
 `VolumeD`           Total session trades of Democratic Party Nominee's 
                     market
 
 `VolumeR`           Total session trades of Republican Party Nominee's 
                    market 
--------------------------------------------------------------------------------

Each row represents daily trading
information about the contracts for either the Democratic or
Republican Party nominee's victory in a particular state. 

We will also
use the election outcome data.  These data files are
`pres08.csv` and `pres12.csv` with variables:

--------------------------------------------------------------------------------
 Name                 Description
 -------------------- ----------------------------------------------------------
 `state.name`         Full name of state (only in `pres2008`)
 
 `state`              Two letter state abbreviation
 
 `Obama`              Vote percentage for Obama
 
 `McCain`             Vote percentage for McCain
 
 `EV`                 Number of electoral college votes for this state
--------------------------------------------------------------------------------
 
We'll also use poll data from 2008 and 2012 in the files `polls08.csv` and `polls12.csv`,  The variables in the polling data are:
  
--------------------------------------------------------------------------------
 Name                 Description
 -------------------- ---------------------------------------------------------
 `state`              Abbreviated name of state in which poll was conducted 
 
 `Obama`              Predicted support for Obama (percentage)
 
 `Romney`             Predicted support for Romney (percentage) 
 
 `Pollster`           Name of organization conducting poll 
 
 `middate`            Middle of the period when poll was conducted 
--------------------------------------------------------------------------------
```{r}
intrade08 <- read.csv("/Users/clayparham/Documents/Bush631-/Assignment 7/intrade08.csv")
intrade12 <- read.csv("/Users/clayparham/Documents/Bush631-/Assignment 7/intrade12.csv")
polls08 <- read.csv("/Users/clayparham/Documents/Bush631-/Assignment 7/polls08.csv")
polls12 <- read.csv("/Users/clayparham/Documents/Bush631-/Assignment 7/polls12.csv")
pres08 <- read.csv("/Users/clayparham/Documents/Bush631-/Assignment 7/pres08.csv")
pres12 <- read.csv("/Users/clayparham/Documents/Bush631-/Assignment 7/pres12.csv")
```
## Question 1

We will begin by using the market prices on the day before the election to predict the 2008 election outcome.  To do this, subset the data such that it contains the market information for each state and candidate only on the day before the election.  Note that in 2008 the election day was November 4.  

We compare the closing prices for the two candidates in a given state and classify a candidate whose contract has a higher price as the predicted winner of that state.  Which states were misclassified?  How does this compare to the classification by polls presented earlier in this chapter? 

Repeat the same analysis for the 2012 election, which was held on November 6.  How well did the prediction market do in 2012 compared to 2008?  Note that in 2012 some less competitive states have missing data on the day before the election because there were no trades on the Republican and Democratic betting markets. Assume Intrade predictions would have been accurate for these states.

## Answer 1


```{r}
intresults08 <- merge(intrade08, pres08, by = "state")
intresults08$day <- as.Date(intresults08$day)
lastdayintrade <-subset(intresults08, subset = (day == "2008-11-03"))
n <- 51
resultsbet <- rep(NA, n)
for (i in 1:n) {
  resultsbet[i] <- lastdayintrade$PriceD[i] - lastdayintrade$PriceR[i]
  lastdayintrade$resultsbet[i] <- resultsbet[i]
  }

resultspoll <- rep(NA, n)
for (i in 1:n) {
  resultspoll[i] <-lastdayintrade$Obama[i] - lastdayintrade$McCain[i]
  lastdayintrade$resultspoll[i]<-resultspoll[i]
}

lastdayintrade$accurate <- ifelse((sign(lastdayintrade$resultspoll) == sign(lastdayintrade$resultsbet)), 1, 0)
mean(lastdayintrade$accurate)


for (i in 1:n) {
  if (lastdayintrade$accurate[i] == 0) {
    cat(lastdayintrade$statename[i], "\n")
  }
}

```



```{r}
intresults12 <- merge(intrade12, pres12, by = "state")
intresults12$day <- as.Date(intresults12$day)
lastdayintrade <-subset(intresults12, subset = (day == "2012-11-03"))
n <- 50
head(lastdayintrade)
resultsbet <- rep(NA, n)


for (i in 1:n) {
  resultsbet[i] <- lastdayintrade$PriceD[i] - lastdayintrade$PriceR[i]
  lastdayintrade$resultsbet[i] <- resultsbet[i]
  
  }

resultspoll <- rep(NA, n)
for (i in 1:n) {
  resultspoll[i] <-lastdayintrade$Obama[i] - lastdayintrade$Romney[i]
  lastdayintrade$resultspoll[i]<-resultspoll[i]
}

for (i in 1:n){
  if (is.na(lastdayintrade$resultsbet[i]) == TRUE){
    lastdayintrade$resultsbet[i] <- lastdayintrade$resultspoll[i]
  }
}

lastdayintrade$accurate <- ifelse((sign(lastdayintrade$resultspoll) == sign(lastdayintrade$resultsbet)), 1, 0)
mean(lastdayintrade$accurate)


for (i in 1:n) {
  if (lastdayintrade$accurate[i] == 0) {
    cat(lastdayintrade$statename[i], "\n")
  }
}
```


## Question 2

How do the predictions based on the betting markets change over time?  

Implement the same classification procedure as above on each of the last 90 days of the 2008 campaign rather than just the day before the election.  

Plot the predicted number of electoral votes for the Democratic party nominee over this 90-day period.  

The resulting plot should also indicate the actual election result. 

Note that in 2008, Obama won 365 electoral votes. Briefly comment on the plot.

## Answer 2
```{r}
year_2008 <- merge(intrade08, pres08, by = "state")

year_2012 <- merge(intrade12, pres12, by = "state")


##what you did here is you created a new data frame
day_before_2008 <- subset(year_2008,
                          subset = (day == "2008-11-03"))

day_before_2012 <- subset(year_2012,
                          subset = (day == "2012-11-05"))



year_2008 $DaysToElection <- as.Date("2008-11-04")  -  as.Date(year_2008 $day)

year_2008 $obama.intmarg <- year_2008 $PriceD  -   year_2008 $PriceR

year_2008 $obama.actmarg <- year_2008 $Obama  -  year_2008 $McCain

latest08 <- year_2008[year_2008 $DaysToElection == 1,]

latest08$state[sign(latest08$obama.intmarg) !=  sign(latest08$obama.actmarg)]


Obama.dailypredEV08 <- rep(NA, 90)

for (i in 1:90) {
    daily.data  <- subset(year_2008,   subset =  (DaysToElection == i))
    Obama.dailypredEV08[i] <-  sum(daily.data$EV[daily.data$obama.intmarg > 0])
} 


plot(1:90, Obama.dailypredEV08,
     main = "Daily Predicted Obama Electoral Votes",
     xlab = "Days to Election", ylab = "Predicted Electoral Votes",
     xlim = c(90, -1), ylim = c(260, 380))
points(0, 365, pch = 19)
abline(v = 0)
abline(h = 365, lty = "dashed")
text(80, 370, "Actual Result") 

#I was working with Jacob on this and had a lot of issues with how I was thinking it should be done. Because of that, I have left my code below for you to look at.
```


```{r}
intresults08 <- merge(intrade08, pres08, by = "state")
intresults08$day <- as.Date(intresults08$day)
last90daysintrade <-subset(intresults08, subset = (day >= "2008-08-06" & day<"2008-11-05"))

last90daysintrade$ObamaW <- ifelse((last90daysintrade$Obama) > (last90daysintrade$McCain), 1, 0)

ObamaWin <- subset(last90daysintrade, subset = (ObamaW == 1))
DayEV <- data.frame(unique(ObamaWin$day))

names(DayEV)[1] <- "Days"


DayEV$Days <- DayEV[order(DayEV$Days),]
DayEV$EV <- NA
n <- nrow(ObamaWin)
y <- nrow(DayEV)
counter = 0
for (e in 1:y){
  for (i in 1:n){
    if (DayEV$Days[e] == ObamaWin$day[i]) {
       counter <- counter+ObamaWin$EV[i]
      
    }
  }
  DayEV$EV[e] <- counter
  counter = 0
  
}
DayEV
```
  

## Question 3

Repeat the previous exercise but this time use the seven-day
  *moving-average* price, instead of the daily price, for each
  candidate within a state.  This can be done with a
  loop. For a given day, we take the average of the Session Close
  prices within the past seven days (including that day).  To answer
  this question, we must first compute the seven-day average within
  each state. Next, we sum the electoral votes for the states Obama is
  predicted to win.  Using the `tapply` function will allow us
  to efficiently compute the predicted winner for each state on a
  given day.
  
## Answer 3

```{r}


rollingaverage <- data.frame(1:90)
rollingaverage
for (i in 1:90){
  rollingaverage[i] <- mean(year_2008$EV)
}
rollingaverage
for (i in 1:90) {
    daily.data  <- subset(year_2008,   subset =  (DaysToElection == i))
    Obama.dailypredEV08[i] <-  sum(daily.data$EV[daily.data$obama.intmarg > 0])
} 


plot(1:90, Obama.dailypredEV08,
     main = "Daily Predicted Obama Electoral Votes",
     xlab = "Days to Election", ylab = "Predicted Electoral Votes",
     xlim = c(90, -1), ylim = c(260, 380))
points(0, 365, pch = 19)
abline(v = 0)
abline(h = 365, lty = "dashed")
text(80, 370, "Actual Result") 

```


## Question 4

Create a similar plot for 2008 state-wide poll predictions using
  the data file `polls08.csv`. Notice that polls are not conducted daily
  within each state. Therefore, within a given state for each of the
  last 90 days of the campaign, we compute the average margin of
  victory from the most recent poll(s) conducted. If multiple polls
  occurred on the same day, average these polls. Based on the most
  recent predictions in each state, sum Obama's total number of
  predicted electoral votes. One strategy to answer this question is
  to program two loops - an inner loop with 51 iterations for each
  state and an outer loop with 90 iterations for each day.
  
## Answer 4

```{r}

```





