---
title: "in class_assign3"
output: html_document
---

SUCCESS OF LEADER ASSASSINATION AS A NATURAL EXPERIMENT

One longstanding debate in the study of international relations concerns the question of whether individual political leaders can make a difference. Some emphasize that leaders with different ideologies and personalities can significantly affect the course of a nation. Others argue that political leaders are severely constrained by historical and institutional forces. Did individuals like Hitler, Mao, Roosevelt, and Churchill make a big difference? The difficulty of empirically testing these arguments stems from the fact that the change of leadership is not random and there are many confounding factors to be adjusted for. In this exercise, we consider a natural experiment in which the success or failure of assassination attempts is assumed to be essentially random.7 Each observation of the CSV data set leaders.csv contains information about an assassination attempt.

Table 2.8 presents the names and descriptions of variables in this leader assassination data set. The polity variable represents the so-called polity score from the Polity Project. The Polity Project systematically documents and quantifies the regime types of all countries in the world from 1800. The polity score is a 21-point scale ranging from −10 (hereditary monarchy) to 10 (consolidated democracy). The result variable is a 10-category factor variable describing the result of each assassination attempt.

1. How many assassination attempts are recorded in the data? How many countries
experience at least one leader assassination attempt? (The unique() function, What is the average number of such attempts (per year) among these countries?

```{r}
leaders <- read.csv("/Users/clayparham/Documents/Bush631-/Assignment Three/leaders.csv")
#names(leaders)
cat("Total assassination attempts:", NROW(leaders$country), "\n")
countryattempt <- unique(leaders$country)
cat("How many countries have had one:", NROW(countryattempt), "\n")
cat("Average number per year:", mean(table(leaders$year)))
```

2. Create a new binary variable named success that is equal to 1 if a leader dies from the attack and 0 if the leader survives. Store this new variable as part of the original data frame. What is the overall success rate of leader assassination? Does the result speak to the validity of the assumption that the success of assassination attempts is randomly determined?

I think that it does speak to the assumption that it is randomly determined. I think this is the case, since the 1's and 0's don't seem to have much correlation to any of the other data.
```{r}
leaders$success <- ifelse((leaders$result == 'dies within a day after the attack') | (leaders$result == 'dies between a day and a week') | (leaders$result =='dies, timing unknown') | (leaders$result == 'dies between a week and a month'), 1, 0)
cat("Success rate:", mean(leaders$success) * 100, "%")

```

3. Investigate whether the average polity score over three years prior to an assassination attempt differs on average between successful and failed attempts. Also, examine whether there is any difference in the age of targeted leaders between successful and failed attempts. Briefly interpret the results in light of the validity of the aforementioned assumption.

There is a difference between the polity scores before and after for successful and failed attempts. More striking, though, is the difference in age between successful and failed attempts. This shows that perhaps it is not truely random if an assassination attempt succeeds or not.
```{r}
cat("Average score for successful:", mean(leaders$politybefore[leaders$success == 1]), "\n")
cat("Average score for unsuccessful:", mean(leaders$politybefore[leaders$success == 0]), "\n")

cat("Average age of successful:", mean(leaders$age[leaders$success == 1]), "\n")
cat("Average age of unsuccessful:", mean(leaders$age[leaders$success == 0]), "\n")

```

4. Repeat the same analysis as in the previous question, but this time using the country’s experience of civil and international war. Create a new binary variable in the data frame called warbefore. Code the variable such that it is equal to 1 if a country is in either civil or international war during the three years prior to an assassination attempt. Provide a brief interpretation of the result.

We see that 36.8% of countries were in a war three years before their assassination attempt. This is a remarkably high number compared to the world generally.
```{r}
names(leaders)
leaders$warbefore <- ifelse((leaders$interwarbefore == 1) | (leaders$civilwarbefore == 1), 1, 0)
mean(leaders$warbefore)
```

5. Does successful leader assassination cause democratization? Does successful leader assassination lead countries to war? When analyzing these data, be sure to state your assumptions and provide a brief interpretation of the results.

The average number of wars after an assassination attempt is 27.6%. However, the average number of wars after a successful attempt is 20.4%. This shows that wars are actually less likely after a successful attempt than an unsuccessful one.

However, the polity score actually does improve after a successful assassination attempt (although this doesn't pass my gut check.)
```{r}
leaders$warafter <- ifelse((leaders$interwarafter == 1) | (leaders$civilwarafter == 1), 1, 0)
##mean of wars after an attempt
mean(leaders$warafter)
##mean of polity score after a success
mean(leaders$polityafter[leaders$success == 1]) - mean(leaders$polityafter[leaders$success == 0])
##mean of wars after a success
mean(leaders$warafter[leaders$success == 1])

```

